---
title: "MGT 6203 - Project"
author: "Team 10"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "`r format(Sys.time(), '%c %Z')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Road traffic crashes are a leading cause of death in many countries, especially those with people living in high-population density environments. In this project, due to data availability issues, we will be focusing on the city of New York City in the United States

## Data Description & Pre-processing

This work is a compiled result of cleaning and using the following data sets:
1. Crashes dataset.
2. Vehicles dataset
3. Persons dataset
4. New York city spatial dataset
5. New York city weather dataset

The cleaned dataset has the following columns:

1. **COLLISION_ID** Auto generated ID for the crash
2. **CRASH.DATETIME** Crash Date and Time
3. **TIME_OF_DAY** Time of Day (morning, afternoon, or evening)
4. **STATE** All the data set records in NY
5. **PO_NAME** Postal Name (48 unique values in NY)
6. **ZIPCODE** ZIP Code (193 unique values in NY)
7. **LATITUDE** Latitude of the crash
8. **LONGITUDE** Longitude of the crash
9. **NUMBER.OF.PERSONS.INJURED** Number of Persons Injured in the crash
10. **NUMBER.OF.PERSONS.KILLED** Number of Persons Killed in the crash
11. **NUMBER.OF.PEDESTRIANS.INJURED** Number of Pedestrians Injured in the crash
12. **NUMBER.OF.PEDESTRIANS.KILLED** Number of Pedestrians Killed in the crash
13. **NUMBER.OF.CYCLIST.INJURED** Number of Cyclist Injured in the crash
14. **NUMBER.OF.CYCLIST.KILLED** Number of Cyclist Killed in the crash
15. **NUMBER.OF.MOTORIST.INJURED** Number of Motorist Injured in the crash
16. **NUMBER.OF.MOTORIST.KILLED** Number of Motorist Killed in the crash
17. **VEHICLE.CATEGORY** 17 vehicle category
17. **POPULATION** Population of the crash area
18. **AREA** Area size where the crash happened
19. **COUNTY** County where the crash happened in NYC (5 Unique values)


```{r}
# Clear variables in memory
rm(list=ls())

# Load required libraries
library(zoo)
library(dplyr)
library(ggplot2)
library(lubridate)


# Reading the data
data <- read.csv("~/Majed AlOtaibi/University/Master/Courses/MGT6203/Project/Data/crashes-cleaned.csv", header=TRUE)
```

```{r}
# summary of the data
summary(data)
str(data)
```
## Varibale Exploration & Selection

```{r}
# Extract the hours as new feature
data$HOUR <- substr(data$CRASH.DATETIME, 12, 13)

# Convert to categorical variables
data$TIME_OF_DAY <- as.factor(data$TIME_OF_DAY)
data$PO_NAME <- as.factor(data$PO_NAME) 
data$ZIPCODE <- as.factor(data$ZIPCODE) 
data$VEHICLE.CATEGORY <- as.factor(data$VEHICLE.CATEGORY) 
data$COUNTY <- as.factor(data$COUNTY)
data$HOUR <- as.factor(data$HOUR)
#"%Y-%m-%d %H:%M:%S"
data$CRASH_DATE <- strptime(as.character(data$CRASH.DATETIME), "%Y-%m-%d")
data$CRASH_DATETIME <- strptime(as.character(data$CRASH.DATETIME), "%Y-%m-%d %H:%M:%S")


# Remove unneeded columns
data$COLLISION_ID <- NULL
data$STATE <- NULL
data$CRASH.DATETIME <- NULL
data$LATITUDE <- NULL
data$LONGITUDE <- NULL
```


```{r}
# summary of the data
summary(data)
str(data)
```


## Split data

Split the data into training and testing data sets

```{r}
# Split data for training and testing
testRows <- sample(nrow(data),0.2*nrow(data))
testData <- data[testRows, ]
trainData <- data[-testRows, ]
```


```{r}
daily_per_injured <- data %>% group_by(CRASH_DATE) %>%
  summarise(daily_per_injured=sum(NUMBER.OF.PERSONS.INJURED, na.rm=TRUE)) %>%
  filter(CRASH_DATE >= "2013-01-01" & CRASH_DATE <= "2022-12-31")
#, sum(NUMBER.OF.PERSONS.KILLED), sum(NUMBER.OF.PEDESTRIANS.INJURED), sum(NUMBER.OF.PEDESTRIANS.KILLED), sum(NUMBER.OF.CYCLIST.INJURED),  sum(NUMBER.OF.CYCLIST.KILLED), sum(NUMBER.OF.MOTORIST.INJURED), sum=(NUMBER.OF.MOTORIST.KILLED)
daily_per_injured[is.na(daily_per_injured)]

#convert data frame to time series
ts_zoo <- read.zoo(daily_per_injured)


ts_monthly <- aggregate(ts_zoo, as.yearmon, sum)
ts_monthly <- ts_monthly[-1,]


ts <- ts(ts_monthly, start=c(2013,1), end=c(2022,12), frequency=12)

plot(ts)

ts_zoo_2013 <- window(ts_zoo, start = as.POSIXlt("2013-01-01 00:00:00"), end = as.POSIXlt("2013-12-31 00:00:00"))

ts_2013_monthly <- aggregate(ts_zoo_2013, as.yearmon, sum)
ts_2013_monthly <- ts_2013_monthly[-1,]

ts_2013 <- ts(ts_2013_monthly, start=c(2013,1), frequency=12)

d <- decompose(ts_2013)

plot(d)

model <- HoltWinters(ts_2013, seasonal="additive")

plot(model)
```

# Trend & Seasonality analysis

```{r}
monthly_data <- data %>%
    filter(CRASH_DATE >= "2013-01-01" & CRASH_DATE <= "2022-12-31") %>%
  group_by(month = factor(as.character(month(CRASH_DATE, label = T, locale = Sys.setlocale("LC_TIME", "English"))),
                            levels = c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec')),
           year = year(CRASH_DATE)) %>%
  summarise(monthly_per_injured=sum(NUMBER.OF.PERSONS.INJURED, na.rm=TRUE),
            monthly_per_killed=sum(NUMBER.OF.PERSONS.KILLED, na.rm=TRUE),
            monthly_ped_injured=sum(NUMBER.OF.PEDESTRIANS.INJURED, na.rm=TRUE),
            monthly_ped_killed=sum(NUMBER.OF.PEDESTRIANS.KILLED, na.rm=TRUE),
            monthly_cyc_injured=sum(NUMBER.OF.CYCLIST.INJURED, na.rm=TRUE),
            monthly_cyc_killed=sum(NUMBER.OF.CYCLIST.KILLED, na.rm=TRUE),
            monthly_mot_injured=sum(NUMBER.OF.MOTORIST.INJURED, na.rm=TRUE),
            monthly_mot_killed=sum(NUMBER.OF.MOTORIST.KILLED, na.rm=TRUE),.groups = "keep")

# Plot monthly persons injured
ggplot(monthly_data, aes(x=month, y=monthly_per_injured, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Persons Injured')   +
xlab('Month') + 
theme_light() +
labs(title="Total Persons Injured per month" ,color = "year")

# Plot monthly persons killed
ggplot(monthly_data, aes(x=month, y=monthly_per_killed, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Persons Killed')   +
xlab('Month') + 
theme_light() +
labs(title="Total Persons Killed per month" ,color = "year")








# Plot monthly Pedestrians injured
ggplot(monthly_data, aes(x=month, y=monthly_ped_injured, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Pedestrians Injured')   +
xlab('Month') + 
theme_light() +
labs(title="Total Pedestrians Injured per month" ,color = "year")

# Plot monthly Pedestrians killed
ggplot(monthly_data, aes(x=month, y=monthly_ped_killed, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Pedestrians Killed')   +
xlab('Month') + 
theme_light() +
labs(title="Total Pedestrians Killed per month" ,color = "year")



# Plot monthly Cyclist injured
ggplot(monthly_data, aes(x=month, y=monthly_cyc_injured, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Cyclist Injured')   +
xlab('Month') + 
theme_light() +
labs(title="Total Cyclist Injured per month" ,color = "year")

# Plot monthly Cyclist killed
ggplot(monthly_data, aes(x=month, y=monthly_cyc_killed, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Cyclist Killed')   +
xlab('Month') + 
theme_light() +
labs(title="Total Cyclist Killed per month" ,color = "year")


# Plot monthly monthly_mot_injured injured
ggplot(monthly_data, aes(x=month, y=monthly_mot_injured, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Motorist Injured')   +
xlab('Month') + 
theme_light() +
labs(title="Total Motorist Injured per month" ,color = "year")

# Plot monthly Motorist killed
ggplot(monthly_data, aes(x=month, y=monthly_mot_killed, color=as.factor(year), group = year)) +
geom_line()   + 
  scale_color_manual(values=c("#FF0000", "#00FF00", "#0000FF","#FFFF00","#00FFFF","#FF00FF","#800080","#008000","#808000","#008080"))+
  theme(legend.position="top")+
ylab('Total Motorist Killed')   +
xlab('Month') + 
theme_light() +
labs(title="Total Motorist Killed per month" ,color = "year")
#rm(list = ls())

#data <- read.delim("~/Majed AlOtaibi/University/Master/Courses/ISYE6501/hw4/temps.txt")

#t <-  as.vector(unlist(data[,2:21]))

#ts <- ts(t, start=1996, frequency = 123)

#ts

#plot(ts)

#model <- HoltWinters(ts, seasonal="multiplicative")

#model$alpha
#model$beta
#model$gamma
#model$coefficients

#plot(model)

#m_seasonal <- matrix(model$fitted[,4], nrow = 123)

#se_data <- as.data.frame(m_seasonal) 

#View(se_data)
```

